---
title: "Weather Event Impact on Human Life and Property from 1996 to 2011"
author: "Becca Smith"
date: "October 16, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Synopsis
The goal of this analysis document is to explore the National Oceanic Atmospheric Administration's (NOAA) storm event data from 1996 to 2011 to discover:  
1. Across the United States, which types of events (as indicated in the EVTYPE variable) were the most harmful with respect to population health? and  
2. Across the United States, which types of events had the greatest economic consequences?

This analysis presents findings showing that excessive heat and tornados caused the most fatalities from 
1996 to 2011, while tornados, floods, and excessive heat caused the most injuries.  The largest economic damage was caused by floods, with hurricanes/ 
typhoons and storm surges/ tide damage also causing large amounts of damage.

# Data Processing
I obtained the National Oceanic Atmospheric Administration's (NOAA) storm event data beginning in 1950 and ending in November 2011.

```{r obtainingdata}
# See if data file is in working directory.  If not, get it
if(!file.exists("repdata%2Fdata%2FStormData.csv.bz2")) {
    fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
    download.file(fileURL, "repdata%2Fdata%2FStormData.csv.bz2")
}
```

Next I read the data into R using read.csv.
```{r readingdata, cache=TRUE}
# See if data read into R.  If not, read it.
if(!exists('stormData')){
    # read data
    stormData <- read.csv("repdata%2Fdata%2FStormData.csv.bz2", header = TRUE,
                          stringsAsFactors = FALSE, na.strings = "NA")
}
```

## Data Transformations (with description and justification)
This is a large dataset.  Here are the dimensions: 
```{r dimstormData}
dim(stormData)
```

To make it easier to work with, I removed rows that do not contain any property damage, crop damage, fatalities, or injuries and limited the columns to those most relevant.
```{r removeharmless, cache=TRUE}
# create dataframe with only events that had health or damage consequence
df <- subset(stormData, (rowSums(
    stormData[, c("FATALITIES", "INJURIES", "CROPDMG","PROPDMG")]))>0,
    select = c(BGN_DATE, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, 
               CROPDMG, CROPDMGEXP, REMARKS, REFNUM))
dim(df)
```

According to NOAA records, information regarding all event types was only tracked starting in 1/1996.  Earlier than that only tornado and a few select other event types were recorded (http://www.ncdc.noaa.gov/stormevents/details.jsp?type=eventtype).  Since I want to compare all event types I removed the events from earlier than  1996. 
```{r after1996, cache=TRUE}
# Start by converting the BGN_DATE column to a date format
df$BGN_DATE <- as.Date(df$BGN_DATE, "%m/%d/%Y")
# Then pull just the year from the date and format as numeric
df$BGN_DATE <- as.numeric(format(df$BGN_DATE,"%Y"))
# Now we can subset the dataframe to rows with a date equal to or greater than
# 1996
df <- subset(df, BGN_DATE >= 1996)
dim(df)
```
# Results
** Instructions **
There should be a section titled Results in which your results are presented.

# Analysis
The analysis document must have at least one figure containing a plot.
Your analysis must have no more than three figures. Figures may have multiple plots in them (i.e. panel plots), but there cannot be more than three figures total.